---
title: <span style="color:#005857"> LPS25 Demo - Biodiversity Data Cubes for Earth Science (22 June, 2025)
subtitle: <span style="color:#47A52A"> Part 3 - Hands-on training  on the `ebvcube` R package <br>  Script to create EBVCubes from tif files
author: 
  - name: Lina Estupinan-Suarez, Luise Quoß
    affiliation: "German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig"
    email: lina.estupinans@idiv.de
institute: "**Institute**: German Centre for Integrative Biodiversity Research (iDiv)"
date: "`r Sys.Date()`"
output:
 html_notebook:
    highlight: tango
---


### Load Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```

```{r}
# Activate to clean your envirionment
rm(list=ls())
gc()
```

```{r}
# Load packages
library(ebvcube)
library(terra)
library(jsonlite)
library(here)
```

```{r}
# Check if working directory point to this code-folder
getwd()
# If not, run the following line of code
# This only runs if you are using RStudio, else use setwd("your/path/to/ebv-workshop2024-main/code")
# setwd(dirname(rstudioapi::getSourceEditorContext()$path))
```


### 1. Prepare all inputs ----

Define paths for reading input data and saving the results.
```{r}
# Path to EBV metadata as JSON
json <- here("part3/data/json/example_ias.json") #if error: add one dot and try again
# json_data <- fromJSON(txt=json) # in case you want to read the jsonfile

# Path to tiffs
pathin <- here("part3/data/tif/input/ias_5sps") #if error: add one dot and try again

# Path to the EBVCube output
nc <- here("part3/data/ebvcubes/ebvcube_example_ias.nc") #if error: add one dot and try again

# Path to the taxonomy file
tax <- file.path(here("part3/data/taxonomy/example_ias_5species.csv"))
```

Load species taxonomy.

```{r}
# Species key to sort the data
# spskey <- names(r1)
spstax <- read.csv(tax)
spstax$species_key <- paste0(10000, 1:nrow(spstax)) # create a key for the species


# # Check whether the spskey order is the same
# spstax$species == spskey
```


### 2. Plot data ----
```{r}
# tifs <- c('entity_1.tif', 'entity_2.tif', 'entity_3.tif') # 3 entities for one metric
tifs <- c('01_ias_total_occurrences.tif', '02_ias_earliest_date_records.tif', '03_ias_latest_date_records.tif') # 3 metrics
tif_paths <- file.path(here("part3/data/tif/input/ias_5sps"), tifs)

# Load the first tif to check the data
rin <- rast(here(tif_paths[1])) # Load only the first band for plotting
spskey <- names(rin)

# Plot the first tif
plot(rin)

spstax$species_key <- spskey
```

### 3. Create EBVCube netCDF ----
Define raster settings and create the EBVCube structure. This netCDF has no data. The data will be added in the next step.

```{r}
# Defining the fillvalue - optional
fillvalue <- -3.4e+38

# Define geographical extent and coordinate system
extent <- c(ext(rin)[1], ext(rin)[2], ext(rin)[3], ext(rin)[4])
res <- res(rin)
epsg <- 3035
                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      fillvalue <- NaN
# Define the precision of the data
prec <- 'integer'
sep <- ','
```

Create the empty EBVCube structure.
```{r}
# Define entity-names
# entities <- c('01_ias_total_occurrences', '02_ias_earliest_date_records', '03_ias_latest_date_records')

# Create empty file (no data yet)
ebv_create_taxonomy(jsonpath = json,
           outputpath = nc,
           taxonomy = tax,
           epsg = epsg,
           extent = extent,
           resolution = res,
           fillvalue = fillvalue,
           prec = prec,
           sep = sep,
           overwrite = TRUE,
           verbose = FALSE # set to TRUE to see additional messages
           )

# Check out the (still empty) datacubes that are available
dc_new <- ebv_datacubepaths(nc, verbose=FALSE)
print(dc_new)
```


### 4. Check metadata
```{r}
# Check data
ebv_datacubepaths(nc)
# ebv_properties(nc, metric=1)
```


```{r}
# Check the metadata of the newly created EBVCube
prop <- ebv_properties(nc, metric=3, verbose=FALSE)

# Print the properties of the EBVCube
print(prop@general)

# Print the spatial information
print(prop@spatial)

# Print the temporal information
print(prop@temporal)
```


### 4. Add data to the (empty) netCDF ----

```{r}
# Add data -----
# Get entity names
entity_names <- ebv_properties(nc, verbose=F)@general$entity_names
```

Map input data to the empty NetCDF
```{r}
for(i in 1:length(spskey)){
print(i)
spx <- which(spskey == spstax[["species_key"]][i])
  ebv_add_data(filepath_nc = nc,
               data = terra::as.matrix(rast(here(tif_paths[1]))[[i]], wide = T),
               metric = 1,
               entity = i,
               timestep = 1,
               ignore_RAM = T)

  ebv_add_data(filepath_nc = nc,
               data = terra::as.matrix(rast(here(tif_paths[2]))[[i]], wide = T),
               metric = 2,
               entity = i,
               timestep = 1,
               ignore_RAM = T)

  ebv_add_data(filepath_nc = nc,
               data = terra::as.matrix(rast(here(tif_paths[3]))[[i]], wide = T),
               metric = 3,
               entity = i,
               timestep = 1,
               ignore_RAM = T)
}
```


### 5. Check your recently created EBVCube
```{r}
# Load the new cube
new_cubepaths <- ebv_datacubepaths(filepath = nc)

mycube <- ebv_read(filepath = nc,
                  datacubepath = new_cubepaths[1,1],
                  entity = 1,
                  type = 'a'
                   )
```    
